---
title: "egssimtools: an R package for the analysis of a model of speciation"
output:
  pdf_document: default
  html_document:
    keep_md: yes
vignette: >
  %\VignetteIndexEntry{vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This package comes with the [ExplicitGenomeSpeciation](https://github.com/rscherrer/ExplicitGenomeSpeciation) program, an individual-based simulation of a speciation event with explicitly modelled diploid genomes and a flexible genotype-phenotype map. This package includes tools to facilitate the extraction and processing of the data generated by the simulation model. Throughout we will use example simulations from the `inst/extdata` folder. 

## Installation

To install the package, click on the RStudio project `egssimtools.Rproj` to open it in RStudio, then in the "Build" tab, click "Install and Restart". The package should be installed on your machine.

Let us now load the package.

```{r, message = FALSE}
library(egssimtools)
library(tidyverse)
library(cowplot)
```

Each simulation is stored in a simulation folder. The content of a simulation folder depends on the settings used when running the program (see detailed [description](https://github.com/rscherrer/ExplicitGenomeSpeciation)). This package introduces functions to retrieve data, but also parameters and genetic architectures saved during the simulations.
